Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-02-08 papers/LOPe/coarray-comparison.tex
Beta-version of text.

We emphasize some of these advantages by comparing the \texttt{Laplace} implementation with the
implementation of the same algorithm from the original Numrich and Reid paper first describing
coarrays in Fortran.  We should point out that this comparison is somewhat unfair, because Numrich
and Reid were introducing coarray notation for transferring memory on distributed memory
architectures, not demonstrating how ideally one should use coarrays within a large application.
%%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
However this example serves
to highlight some of the advantages of LOPe and CAFe as introduced above.  Note that in the coarray
example shown below, type declarations have been removed to save space:
{\small \begin{verbatim}
  subroutine Laplace (nrow,ncol,U)
    left = me-1      ! me refers to the current image
    if (me == 1) left = ncol
    right = me + 1
    if (me == ncol) right = 1
    call sync_all( [left,right] )   ! Wait for left and right
    new_u(1:nrow) = new_u(1:nrow) + u(1:nrow)[left] + u(1:nrow)[right]
    call sync_all( [left,right] )
    u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
  end subroutine
\end{verbatim}}

The advantages of LOPe and CAFe over this example are now described.  Please note that the following
is not a criticism of coarray Fortran as CAF is a general purpose parallel programming language and
LOPe only pertains to the halo pattern most useful in the implementation of stencil-based algorithms.

\subsubsection{LOPe advantages.}
\begin{itemize}

@@

\item
Algorithm implementation is separate from user-specified synchronization, e.g., \texttt{call
  sync\_all}.  In LOPe, synchronization is subsummed in the semantics of the \texttt{CONCURRENT}
attribute and the \texttt{TRANSFER\_HALO} function call.

@@

\item
The algorithm implementation is separate from any specification as to where the algorithm
is to be executed.  The CAF example explicity denotes where a statement is to executed
with control flow construct like \texttt{if (me == 1)}.

@@

\subsubsection{Errors that are constrained by the language.}
\begin{itemize}

\item
A programmer is not able to store data to the halo region during execution of a LOPe concurrent
function.  If this were allowed, one thread could overwrite another threads data at undefined times.
The compiler is able to catch this class of error.

\item
A programmer can't make indexing errors in a concurrent routine by going out of bounds of the array
plus halo memory.  The compiler is able to catch this class of error at compile time as long as
compile-time constants are used to specify halo sizes.

\item
A programmer is not able to cause race conditions by forgetting to create and use temporary arrays
properly.  In LOPe it is the compilers responsibility to store data in temporary memory.

\item
A programmer can't make synchronization errors as synchronization is implicit in the
\texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided with a copy of
it's local array element plus halo that is consistent with the state of memory \emph{at the time of
  invocation of the procedure}.  Stores to an individual thread's local array element (by that
thread) is never visible to other threads.  LOPE encourages the creation of small functions and lets
the compiler fuse the functions together to improve performance and to provide necessary
synchronization.

>>>>

We emphasize some of these advantages by comparing the \texttt{Laplace} implementation with the
implementation of the same algorithm from the original Numrich and Reid
paper\cite{Numrich:1998:CFP:289918.289920} first describing coarrays in Fortran.  We should point
out that this comparison is somewhat unfair, because Numrich and Reid were introducing coarray
notation for transferring memory on distributed memory architectures, not demonstrating how ideally
one should use coarrays within a large application.

%%Please note that this example is somewhat unfair because in practice CAF is usually refactored in a locally oriented way by so that communication and synchronization separated into separate procedures.  Locally-oriented programming should be viewed as programming methodology with LOPe as a particular instance.
However this example serves
to highlight some of the advantages of LOPe and CAFe as introduced above.  Note that in the coarray
example shown below, type declarations have been removed to save space:
{\small \begin{verbatim}
  subroutine Laplace (nrow,ncol,U)
    left = me-1      ! me refers to the current image
    if (me == 1) left = ncol
    right = me + 1
    if (me == ncol) right = 1
    call sync all    ! Wait for left and right
    new_u(1:nrow) = new_u(1:nrow) + u(1:nrow)[left] + u(1:nrow)[right]
    call sync all
    u(1:nrow) = new_u(1:nrow) - 4.0*u(1:nrow)
  end subroutine
\end{verbatim}}

%%The advantages of LOPe and CAFe over this example are now described.  Please note that the following is not a criticism of coarray Fortran as CAF is a general purpose parallel programming language and LOPe only pertains to the halo pattern most useful in the implementation of stencil-based algorithms.

\subsection{LOPe Advantages}

\begin{itemize}

@@

\item
Algorithm implementation is separate from user-specified synchronization, e.g., \texttt{call
  sync\_all}.  In LOPe, synchronization is subsumed in the semantics of the \texttt{CONCURRENT}
attribute and the \texttt{TRANSFER\_HALO} function call.

@@

\item
The algorithm implementation is separate from any specification as to where the algorithm
is to be executed.  The CAF example explicity denotes where a statement is to executed
with a control flow construct like \texttt{if (me == 1)}.

@@

\subsection{LOPe Constraints}

Contraints provided by the LOPe language extensions allow the compiler to catch several classes of
errors that otherwise would be the programmer's responsibility:

\begin{itemize}

\item
A programmer is not able to store data to the halo region during execution of a LOPe concurrent
function.  Neither are stores to the local element, followed by a read from the halo region of the
same variable allowed.  If these were allowed, one thread could overwrite another threads data at
undefined times.

\item
A programmer can't make indexing errors in a concurrent routine by going out of bounds of the array
plus halo memory.

\item
A programmer is not able to cause race conditions by forgetting to create and use temporary arrays
properly.

\item
A programmer can't make synchronization errors in calls to LOPe functions as synchronization is
implicit in the \texttt{CONCURRENT} attribute.  A thread running a concurrent procedure is provided
with a copy of it's local array element plus halo that is consistent with the state of memory
\emph{at the time of invocation of the procedure}.  Stores to an individual thread's local array
element (by that thread) are never visible to other threads.  Normal coarray programs (like MPI)
require explicit synchronization to ensure that processes arrive at the same program location
before a memory read occurs (for example).  LOPE encourages the creation of small functions and
lets the compiler fuse the functions together to improve performance and to provide necessary
synchronization.

### 2015-02-08 papers/LOPe/coarray-extensions.tex
Beta-version of text.

Consider the \texttt{Laplace} concurrent function discussed above.  In
this section we demonstrate how this function can be called in the
normal context of a program, one that allows full access to all of the
interior elements of the array, as well as array elements within the
logically exterior, halo-boundary region.  Topics highlighted in this
section are: 1. Distributed memory array allocation; 2. Explicit
memory placement; 3. Remote memory transfer; and 4. Remote execution.
This description is within the context of an extension of the Fortran
2008 language.  As shorthand, these specific extensions are referred
to as CAFe, for Coarray Fortran extensions.

@@

We start with the declaration of an array (to be allocated later),
\begin{verbatim}
  real, allocatable, HALO(1:*:1,1:*:1),          &
        dimension(:,:), codimension[:,:] :: U
\end{verbatim}
with an explicit halo size and with two local dimensions (rank) and
two distributed memory codimensions (corank).  The corank of the array
is chosen to be identical to the rank of the array so that the logical
process topology aligns in a way that allows a natural halo exchange
between logically neighboring processes.\todo{What if corank and rank are not the same?}  For example, if the process
location is \texttt{[pcol,prow]}, then the right-hand halo for the
local array \texttt{U} can be obtained by the assignment
\texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size and cosize of
\texttt{U} are given by the allocation statement,
\texttt{allocate(U(0:M+1,0:N+1)[MP,*])}.
This allocation statement specifies (given the one element halo size
provided earlier for \texttt{U}) that the left halo column is
\texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the bottom row
is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the
interior region \texttt{U(1:M,1:N)}.  In this allocation statement,
the total number of process rows \texttt{NP} can be obtained at
runtime, but \emph{may not} be explicitly provided (according to CAF
rules) because the actual number of participating processes (in
Fortran called images) is variable, depending on how many processes
are requested at program startup.  In this discussion, it is
\emph{assumed} that there are no holes in the logical processor
topology, thus $MP*NP = P$, where $P$ is the total number of
participating processes (images).

So far --- except for the \texttt{HALO} attribute --- this is totally
standard Fortran (2008).  For CAFe, we extend the concept of Fortran
images to be hierachical.  By this we mean that each image \emph{may}
have a subimage (or subimages), but this subimage is not visible to
other regular Fortran images.  Subimages also execute differently than
normal images.  Subimages are task based while images all execute a
Single Program but with different (Multiple) Data (SPMD).  A task can
be given to a subimage, but execution on the subimage terminates once
the task is finished.  Memory on a subimage is permanent, however, and
must be specifically allocated and deallocated.

One obtains a subimage by executing a function call,
\texttt{device = GET\_SUBIMAGE(1)}
where the integer argument representing one of attached hardware
devices\todo{mention the relation of subimages and hardware earlier in
  defn. of subimage}.  If the function fails (e.g., the requested
device is unavailable) it returns the image number
\texttt{this\_image()} of the process that is executing the current
program.  Returning the current image allows program execution to
proceed correctly even if there are no attached devices.

Once a subimage is obtained, memory on the device may be allocated,
\begin{verbatim}
  if (device /= this_image()) then
    allocate(U[device], HALO_SRC=U)   [[device]]
  end if
\end{verbatim}
There are four points to note regarding this memory allocation: 1. Memory is only allocated if a
subimage has been obtained; 2. The location where memory is allocated is denoted by regular coarray
notation \texttt{U[device]}; 3. The allocated size and halo attribute of the new array are assumed
from the previously allocated local array \texttt{U} via the notation \texttt{HALO\_SRC=U} (using
\texttt{HALO\_SRC} will also initially copy \texttt{U} to the subimage); and finally 4. The
allocation itself is \emph{executed} on the subimage device with the notation \texttt{[[device]]}.

@@

Execution of the \texttt{Laplace} task is done using the \texttt{do}
\texttt{concurrent} construct:
\begin{verbatim}
   do while (.not. converged)
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplace( U(i,j)[device] )
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do
\end{verbatim}
There are several points that require highlighting: 1. Iteration occurs over the interior
of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop body occurs on the
specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
in any order, even \emph{concurrently}; 4. The local element of the array (as
defined above in reference to the definition of the concurrent procedure
\texttt{Laplace}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
task is to be taken from the subimage as noted by \texttt{[device]}; 5. All threads must finish
execution of the loop body before further execution of the program proceeds; and 6. Transfer of
all requisite halo regions is effected by the call to the new LOPe intrinsic function
\texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must
complete the halo transfer before program execution continues.

Note that a transfer of halo memory is necessary after each completion of the do concurrent loop.
This must be done in order for the halo region of a coarray on a given process to be consistent
with the corresponding interior of the coarray on a logically neighboring process.

%% However, recall that with the LOPe programming model only the local element is mutable; thus, the halo region of an array needs update.

Finally, memory for the entire array \texttt{U} can be copied from the
subimage device with the statement,
\texttt{U = U[device]}, 
and memory deallocation (not shown) is similar to memory allocation.

@@

Note that the use of halo cells is the normal way that large and complex MPI and CAF
programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran
language allowing the compiler access to this information in order to spread computation over more
hardware resources, improve performance, and to reduce complexity for the programmer.

>>>>

Consider the \texttt{Laplacian} concurrent function in Fig. 1.  In this section we demonstrate how
this function can be called in the normal context of a program, one that allows full access to all
of the interior elements of the array, as well as array elements within the logically exterior,
halo-boundary region.  Topics highlighted in this section are: 1. distributed memory array
allocation; 2. explicit memory placement; 3. remote memory transfer; and 4. remote execution.  This
description is within the context of extensions to Fortran; as shorthand, these extensions are
referred to as CAFe, for Coarray Fortran extensions.  CAFe is complementary to previous work
extending coarray Fortran\cite{mellor-crummey:2009:caf2,jin:2011:caf2}.

We introduce the important new concept of a CAFe subimage.  Fortran images are a collection of
distributed memory processes that all run the same program (image).  LOPe extends the concept of a
Fortran image by allowing images to be hierachical.  By this we mean that each image \emph{may}
have a subimage (or subimages), but this subimage is not visible to other regular Fortran images.
Subimages also execute differently than normal images and may execute on different non-homogeneous
hardware, .e.g, an attached accelerator device.  Subimages are task based while images all execute
a Single Program but with different (Multiple) Data (SPMD).  A task can be given to a subimage, but
execution on the subimage terminates once the task is finished.  Memory on a subimage is permanent,
however, and must be specifically allocated and deallocated.

One obtains a subimage by executing the new LOPe function call, \texttt{device = GET\_SUBIMAGE(1)},
where the integer argument represents an attached hardware device (or a separate process).
If the function fails (e.g., the requested device is unavailable) it returns the image number
\texttt{this\_image()} of the process that is executing the current program.  Returning the current
image allows program execution to proceed correctly even if there are no attached devices.

@@

We start with the declaration of an array with an explicit halo size and with two local
dimensions (rank) and two distributed memory codimensions (corank),
\begin{verbatim}
   real, allocatable, dimension(:,:), codimension[:,:]      &
         HALO(1:*:1,1:*:1) :: U
\end{verbatim}
The corank of the array is chosen to be identical to the rank of the array so that the logical
process topology aligns in a way that allows a natural halo exchange between logically neighboring
processes (this could not occur if corank and rank are not the same).  For example, if the process
location is \texttt{[pcol,prow]}, then the right-hand halo for the local array \texttt{U} can be
obtained by the assignment
\texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size and cosize of
\texttt{U} are given by the allocation statement,
\texttt{allocate(U(0:M+1,0:N+1)[MP,*])}.
This allocation statement specifies (given the one element halo size provided earlier for
\texttt{U}) that the left halo column is \texttt{U(0,:)}, the right column is \texttt{U(M+1,:)},
the bottom row is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the interior region
\texttt{U(1:M,1:N)}.

In this allocation statement, the total number of process columns $NP$ can be obtained at runtime,
but \emph{may not} be explicitly provided (according to CAF rules) because the actual number of
participating processes (in Fortran called images) is variable, depending on how many processes are
requested at program startup.  In this discussion, it is \emph{assumed} that there are no holes in
the logical processor topology, thus $MP*NP = P$, where $MP$ is the number of process rows and $P$
is the total number of participating processes (images).

Once a subimage is obtained, memory on the device can be allocated,
\begin{verbatim}
   if (device /= this_image()) then
      allocate(U[device], HALO_SRC=U)   [[device]]
   end if
\end{verbatim}
There are four points to note regarding this memory allocation: 1. Memory is only allocated if a
subimage has been obtained; 2. The location where memory is allocated is denoted by regular coarray
notation \texttt{U[device]}; 3. The allocated size and halo attribute of the new array are obtained
from the previously allocated local array \texttt{U} via the notation \texttt{HALO\_SRC=U} (using
\texttt{HALO\_SRC} will also initially copy \texttt{U} to the subimage); and finally 4. The
allocation itself is \emph{executed} on the subimage device with the notation \texttt{[[device]]}.

@@

Execution of the \texttt{Laplacian} task is done using the \texttt{do}
\texttt{concurrent} construct:
\begin{verbatim}
   do while (.not. converged)
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplacian( U(i,j)[device] )
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do
\end{verbatim}
There are several points that require highlighting: 1. Iteration occurs over the interior
of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop body occurs on the
specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
in any order, even \emph{concurrently}; 4. The local element of the array (as
defined above in reference to the definition of the concurrent procedure
\texttt{Laplacian}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
task is to be taken from the subimage as noted by \texttt{[device]}; 5. All threads must finish
execution of the loop body before further execution of the program proceeds; and 6. Transfer of
all requisite halo regions is effected by the call to the new LOPe intrinsic function
\texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must
complete the halo transfer before program execution continues.

Note that a transfer of halo memory is necessary after each completion of the do concurrent loop.
This must be done in order for the halo region of a coarray on a given process to be consistent
with the corresponding interior of the coarray on a logically neighboring process.
Finally, memory for the entire array \texttt{U} can be copied from the
subimage device with the statement,
\texttt{U = U[device]}, 
and memory deallocation (not shown) is similar to memory allocation.

@@

%
% This could be part of conclusions
%

%%Note that the use of halo cells is the normal way that large and complex MPI and CAF programs are implemented.  LOPe proposes to formalize this common pattern into the Fortran language allowing the compiler access to this information in order to spread computation over more hardware resources, improve performance, and to reduce complexity for the programmer.
